---
title: "Adherence"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(readxl)
library(UpSetR)
library(brms)
library(cmdstanr)
library(bayesplot)
options(mc.cores = 4, brms.backend = "cmdstanr")
cache_dir <- here::here("stored_fits")
if(!dir.exists(cache_dir)) {
  dir.create(cache_dir)
}

run_pp_checks <- TRUE

devtools::load_all()

all_data <- load_data_adherence1()
data_wide <- all_data$wide
data_long <- all_data$long
```

Following subjects have discrepancy between the "Má XX" column and measured levels:

```{r}
all_data$long_raw %>% select(subject_id, drug_class, level, has) %>%
  group_by(subject_id, drug_class, has) %>%
  summarise(has_from_levels = any(!is.na(level)), .groups = "drop") %>%
  filter(has != has_from_levels)
```

```{r}
all_data$long_for_checks %>% 
    select(subject_id, drug_class, adherent, adherent_manual) %>%
    filter(adherent_manual != adherent) %>%  distinct()  

```


```{r}
data_upset <- data_wide %>% 
  select(subject_id, starts_with("has."))
names(data_upset) <- gsub("^has.", "", names(data_upset))
  
upset(as.data.frame(data_upset), sets = names(data_upset)[2:ncol(data_upset)], order.by = "freq", nintersects = NA, nsets = 10)
```

# Exploratory plots

```{r}
data_long %>% filter(has) %>% group_by(drug_class) %>%
  summarise(yes = sum(adherent == "yes"), no = sum(adherent == "no"),
            proportion = yes / (yes + no), 
            low95 = qbeta(0.025, yes + 1, no + 1),
            high95 = qbeta(0.975, yes + 1, no + 1),
            total = yes + no) %>%
  ggplot(aes(x = drug_class, y = proportion, ymin = low95, ymax = high95, label = total)) + 
    geom_linerange() + geom_point(size = 2) + geom_label(y = 1.1) + expand_limits(y = 1.1)
```

```{r}
N_drugs_per_subject <- data_long %>%
  filter(has) %>%
  group_by(subject_id) %>%
  summarise(N_adherent = sum(adherent_int), N_drugs = sum(has)) %>%
  group_by(N_drugs, N_adherent) %>%
  summarise(count = n())

crossing(N_adherent = 0:max(N_drugs_per_subject$N_drugs), N_drugs = 1:max(N_drugs_per_subject$N_drugs)) %>% 
  filter(N_adherent <= N_drugs) %>%
  left_join(N_drugs_per_subject) %>%
  group_by(N_drugs) %>%
  mutate(count = replace_na(count, 0), total = sum(count)) %>%
  group_by(N_drugs, N_adherent, count) %>%
  summarise(
    proportion = count / total, 
            low95 = qbeta(0.025, count + 1, total - count + 1),
            high95 = qbeta(0.975, count + 1, total - count + 1)) %>%
  ggplot(aes(x = N_adherent, y = count)) + 
    geom_line() +
    geom_point(size = 2) +
    facet_wrap(~N_drugs, scales = "free_y", labeller = label_both)
  # ggplot(aes(x = N_adherent, y = proportion, ymin = low95, ymax = high95, label = count)) + 
  #   geom_linerange() + geom_point(size = 2) + geom_label(y = 1.1) + expand_limits(y = 1.1) +
  #   facet_wrap(~N_drugs)
  
  
```

```{r}
data_long %>%
  filter(has) %>%
  group_by(subject_id) %>%
  summarise(N_adherent = sum(adherent_int), N_drugs = sum(has),
            category = case_when(N_adherent == 0 ~ "A - none",
                                 N_adherent == N_drugs ~ "E - all",
                                 N_adherent <= 0.5 * N_drugs ~ "B - half or less",
                                 N_adherent == N_drugs - 1 ~ "D - all but one",
                                 TRUE ~ "C - in between"
                                 )) %>%
  group_by(category) %>%
  summarise(n())
```

Same as above, but excluding furosemide

```{r}
data_long %>%
  filter(has, drug_class != "furosemide") %>%
  group_by(subject_id) %>%
  summarise(N_adherent = sum(adherent_int), N_drugs = sum(has),
            category = case_when(N_adherent == 0 ~ "A - none",
                                 N_adherent == N_drugs ~ "E - all",
                                 N_adherent <= 0.5 * N_drugs ~ "B - half or less",
                                 N_adherent == N_drugs - 1 ~ "D - all but one",
                                 TRUE ~ "C - in between"
                                 )) %>%
  group_by(category) %>%
  summarise(n())
```

## Adherence by various factors

```{r}
plot_adherence_by_continuous <- function(data_long, by_val) {
  data_long %>%
    filter(has) %>%
    group_by(drug_class) %>%
    filter(any(adherent_int == 0)) %>%
    ggplot(aes(y = adherent_int, x = {{by_val}})) + stat_smooth(method = "glm", method.args=list(family="binomial"), formula = y ~ x) + geom_point(position = position_jitter(height = 0.1)) + facet_wrap(~drug_class)
}


plot_adherence_by_discrete <- function(data_long, by_val) {
  data_long %>%
    filter(has, !is.na({{by_val}})) %>%
    group_by(drug_class) %>%
    filter(any(adherent_int == 0)) %>%
    mutate({{by_val}} := as_factor({{by_val}})) %>%
    group_by(drug_class, {{by_val}}) %>%
    summarise(yes = sum(adherent == "yes"), no = sum(adherent == "no"),
            proportion = yes / (yes + no), 
            low95 = qbeta(0.025, yes + 1, no + 1),
            high95 = qbeta(0.975, yes + 1, no + 1),
            total = yes + no, .groups = "drop") %>%
  ggplot(aes(x = {{by_val}}, y = proportion, ymin = low95, ymax = high95, label = total)) + 
    geom_linerange() + geom_point(size = 2) + geom_label(y = 1.2, size = 3) + expand_limits(y = 1.25) +
    facet_wrap(~drug_class)
}


plot_adherence_by_continuous(data_long, age)
plot_adherence_by_discrete(data_long, sex)
plot_adherence_by_continuous(data_long, BMI)
plot_adherence_by_discrete(data_long, NYHA)
plot_adherence_by_continuous(data_long, systolic_BP)
plot_adherence_by_continuous(data_long, diastolic_BP)
plot_adherence_by_continuous(data_long, EF_numeric)
plot_adherence_by_continuous(data_long, pmin(EF_first_amublance_numeric, EF_first_contact_numeric))
plot_adherence_by_continuous(data_long, ekg_freq)
plot_adherence_by_continuous(data_long, lab_Na)
plot_adherence_by_continuous(data_long, lab_Cl)
plot_adherence_by_continuous(data_long, log(lab_U))
plot_adherence_by_continuous(data_long, lab_K)
plot_adherence_by_continuous(data_long, log(lab_Kr))
plot_adherence_by_continuous(data_long, log(lab_NT_proBNP))
plot_adherence_by_continuous(data_long, lab_GFR) + geom_vline(color = "red", xintercept = 1.55)


```
```{r}

```


```{r}
#unique(data_long$VO2_max)
plot_adherence_by_continuous(data_long, as.numeric(VO2_max))

```

## Drugs and their expected metabolic effect

```{r}
data_long %>% filter(drug_class == "BB") %>%
  ggplot(aes(x = adherent, y = ekg_freq)) + geom_boxplot() + geom_point(position = position_jitter(width = 0.3))
```

## Adherence questionnaire

The questionnaire was:

- Q1: Have you ever forgotten to take your heart failure medication. Orig: "Zapomněl jste si někdy vzít léky na srdeční selhání?"
- Q2: Do you dutifully use your heart failure medication? Orig: "Jste pečlivý v užívání léků na srdeční selhání?"
- Q3: When you feel well, do you sometimes skip your heart failure medication? Orig: "Když se cítíte dobře, vysadíte si někdy léky na srdeční selhání?"
- Q4: Do you sometimes skip your heart failure medication when you feel unwell? Orig: "Někdy, když se cítíte špatně, vysadíte si léky na srdeční selhání?"

```{r}


plot_adherence_by_discrete(data_long, questionnaire_1) 
plot_adherence_by_discrete(data_long, questionnaire_2)
plot_adherence_by_discrete(data_long, questionnaire_3)
plot_adherence_by_discrete(data_long, questionnaire_4)
plot_adherence_by_discrete(data_long %>% mutate(q_total = questionnaire_1 - questionnaire_2 + questionnaire_3 + as.numeric(questionnaire_4) ), q_total)

```





# Modelling

```{r}
data_for_model1 <- data_long %>% filter(has) 

if(any(is.na(data_for_model1$adherent_int))) {
  stop("NAs")
}
priors1 <- prior(normal(0,2), class = "sd")
fit1 <- brm(adherent_int ~ (1 | drug_class) + (1 | subject_id), family = "bernoulli", data = data_for_model1, prior = priors1, 
            file = paste0(cache_dir, "/fit1"), file_refit = "on_change", refresh = 500)

summary(fit1)
ranef(fit1)$drug_class
```



```{r}
preds <- posterior_predict(fit1)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$drug_class)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$sex)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = interaction(data_for_model1$sex, data_for_model1$drug_class), freq = FALSE)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = cut(data_for_model1$age, 4))
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = interaction(data_for_model1$sex, cut(data_for_model1$age, 4)))
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$NYHA, freq = FALSE)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = replace_na(data_for_model1$EF, "NA"), freq = FALSE)

```


TODO:
plot observed adherence vs. fit1 predicted adherence

```{r}
fit_simple <- brm(adherent_int ~ (1 | drug_class), family = "bernoulli", data = data_for_model1, prior = priors1, 
            file = paste0(cache_dir, "/fit_simple"), file_refit = "on_change", refresh = 500)

summary(fit_simple)
ranef(fit_simple)$drug_class
```

```{r}
preds <- posterior_predict(fit_simple)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$drug_class)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$sex)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = cut(data_for_model1$age, 4))
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = interaction(data_for_model1$sex, cut(data_for_model1$age, 4)))
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = data_for_model1$NYHA)
ppc_bars_grouped(data_for_model1$adherent_int, yrep = preds, group = replace_na(data_for_model1$EF, "NA"))

```


```{r}
data_aggregate <- data_long %>% filter(has) %>%
  group_by_at(vars(- all_of(c("drug_class", "has", "adherent", "drug_superclass", "adherent_int")))) %>%
  summarise(N_drugs = sum(has), N_adherent = sum(adherent_int), 
            prop = N_adherent / N_drugs,
            .groups = "drop")

if(nrow(data_aggregate) != nrow(data_wide)) {
  stop("Bad processing")
}

if(!identical(sort(unique(data_aggregate$subject_id)), sort(unique(data_wide$subject_id)))) {
  stop("Bad processing")
}
```


```{r}
fit_aggregate_simple <- brm(N_adherent | vint(N_drugs) ~ 1, data = data_aggregate, family = beta_binomial2, stanvars = stanvars_beta_binomial2, refresh = 500, file = paste0(cache_dir, "/fit_aggregate_simple"), file_refit = "on_change")

summary(fit_aggregate_simple)
```

```{r}
preds <- posterior_predict(fit_aggregate_simple)
preds_prop <- sweep(preds, MARGIN = 2, STATS = data_aggregate$N_drugs, FUN = "/")
ppc_stat_grouped(data_aggregate$prop, preds_prop, group = data_aggregate$sex)
ppc_stat_grouped(data_aggregate$prop, preds_prop, group = data_aggregate$sex, stat = sd)

ppc_stat_grouped(data_aggregate$prop, preds_prop, group = cut(data_aggregate$age, 4))
ppc_stat_grouped(data_aggregate$prop, preds_prop, group = cut(data_aggregate$age, 4), stat = sd)


ppc_stat_grouped(data_aggregate$prop, preds_prop, group = data_aggregate$NYHA)
ppc_stat_grouped(data_aggregate$prop, preds_prop, group = data_aggregate$NYHA, stat = sd)

ppc_stat_grouped(data_aggregate$prop, preds_prop, group = substr( replace_na(data_aggregate$EF, "NA"),1, 1)) + ggtitle("EF/10")
ppc_stat_grouped(data_aggregate$prop, preds_prop, group = substr( replace_na(data_aggregate$EF, "NA"),1, 1), stat = sd)  + ggtitle("EF/10")

```


## Bad health



VO2 max, NYHA, EF, NT-proBNP

```{r}
data_for_model1 %>% group_by(EF) %>% summarise(n())
data_wide %>% select(EF, VO2_max, NYHA, lab_NT_proBNP) %>% mutate(EF = as.integer(factor(EF)) + runif(n(), -0.1,0.1), NYHA = as.integer(factor(NYHA))  + runif(n(), -0.1,0.1), lab_NT_proBNP = log(lab_NT_proBNP)) %>% as.matrix() %>% pairs()
```

```{r}
data_for_model1 %>% group_by(subject_id) %>%
  summarise(prop_adherent = sum(adherent_int) /  n()) %>%
  group_by(prop_adherent) %>%
  summarise(n())
```

```{r}
mod_badhealth <- cmdstan_model("badhealth.stan")
```
```{r, results = 'hide'}
subject_id_map <- 1:length(unique(data_for_model1$subject_id))
names(subject_id_map) <- unique(data_for_model1$subject_id)

patient_data <- data_for_model1 %>% select(subject_id,
                                           NYHA,
                                           VO2_max,
                                           EF,
                                           lab_NT_proBNP) %>%
  distinct() %>%
  mutate(subject_id_model = subject_id_map[as.character(subject_id)],
         NYHA = factor(NYHA),
         EF = factor(EF)) %>%
  arrange(subject_id_model)

data_badhealth <- list(
  N = nrow(data_for_model1),
  adherent = data_for_model1$adherent_int,
  N_classes = length(unique(data_for_model1$drug_class)),
  classes = factor(data_for_model1$drug_class),
  N_subjects = length(subject_id_map),
  subject = subject_id_map[as.character(data_for_model1$subject_id)],
  N_NYHA = length(levels(patient_data$NYHA)),
  NYHA = patient_data$NYHA,
  VO2_max_missing = is.na(patient_data$VO2_max),
  VO2_max = replace_na(patient_data$VO2_max, 0),
  NT_proBNP = patient_data$lab_NT_proBNP,
  N_EF = length(levels(patient_data$EF)),
  EF_missing = is.na(patient_data$EF),
  EF = replace_na(as.integer(patient_data$EF), 0)
  
)

cache_file <- file.path(cache_dir, paste0("fit_badhealth_", rlang::hash(data_badhealth), "_", rlang::hash(mod_badhealth$code())))
if(file.exists(cache_file)) {
  fit_badhealth <- readRDS(cache_file)
} else {
  fit_badhealth <- mod_badhealth$sample(data = data_badhealth, refresh = 500)
  fit_badhealth$save_output_files(dir = cache_dir)
}
  
```

```{r}
fit_badhealth$summary(variables = c("badhealth_slope", "nyha_thres", "VO2_max_intercept", "VO2_max_slope","VO2_max_sd",  "NT_proBNP_intercept",
                                    "NT_proBNP_slope", "NT_proBNP_sd", "EF_thres"))

worst_health_model_ids <- fit_badhealth$summary(variables = "badhealth") %>% arrange(desc(q5)) %>% slice_head(n = 5) %>% pull(variable) %>%
  gsub("badhealth\\[|\\]", "", .)
worst_health_subject_ids <- names(subject_id_map[subject_id_map %in% worst_health_model_ids])
```


```{r}
badhealth_rvars <- posterior::as_draws_rvars(fit_badhealth$draws())
linpred_adherent <- badhealth_rvars$adherent_intercept + badhealth_rvars$class_intercept[data_badhealth$classes] + 
  badhealth_rvars$badhealth_slope * badhealth_rvars$badhealth[data_badhealth$subject] 

epred_adherent <- posterior::rdo(plogis(linpred_adherent))

pred_adherent <- posterior::draws_of(posterior::rvar_rng(rbinom, n = data_badhealth$N, prob = epred_adherent, size = 1))
```

```{r}
ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = data_for_model1$sex)

ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = cut(data_for_model1$age, 4))
ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = data_for_model1$NYHA)
ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = replace_na(as.character(cut(data_for_model1$VO2_max, 4)), "NA"))
ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = cut(log(data_for_model1$lab_NT_proBNP), 4))



ppc_bars_grouped(data_badhealth$adherent, pred_adherent, group = replace_na(substr( data_for_model1$EF,1, 1), "NA")) + ggtitle("EF/10")
```
```{r}
linpred_NT_proBNP <- badhealth_rvars$NT_proBNP_intercept + badhealth_rvars$NT_proBNP_slope * badhealth_rvars$badhealth
pred_NT_proBNP <- posterior::draws_of(posterior::rvar_rng(rlnorm, n = data_badhealth$N_subjects,  linpred_NT_proBNP, badhealth_rvars$NT_proBNP_sd))

ppc_dens_overlay(log(data_badhealth$NT_proBNP), log(pred_NT_proBNP[sample(1:4000, 50),]))
ppc_dens_overlay_grouped(log(data_badhealth$NT_proBNP), log(pred_NT_proBNP[sample(1:4000, 50),]), group = fct_collapse(patient_data$NYHA, "3-4" = c("3", "3-4")))
```
```{r}
badhealth_noised <- posterior::rvar_rng(rlogis, n = data_badhealth$N_subjects, location = badhealth_rvars$badhealth)
pred_NYHA <- 1
for(i in 1:dim(badhealth_rvars$nyha_thres)) {
  pred_NYHA <- pred_NYHA + (badhealth_noised > badhealth_rvars$nyha_thres[i])
}

pred_NYHA <- posterior::draws_of(pred_NYHA)
obs_NYHA <- as.integer(data_badhealth$NYHA)

ppc_bars(obs_NYHA, pred_NYHA)
ppc_bars_grouped(obs_NYHA, pred_NYHA, group = cut(log(patient_data$lab_NT_proBNP), 4))
```

```{r}
pred_EF <- 1
for(i in 1:dim(badhealth_rvars$EF_thres)) {
  pred_EF <- pred_EF + (badhealth_noised > badhealth_rvars$EF_thres[i])
}

obs_EF <- as.integer(data_badhealth$EF)
pred_EF <- posterior::draws_of(pred_EF)[,!data_badhealth$EF_missing]
obs_EF <- obs_EF[!data_badhealth$EF_missing]

ppc_bars(obs_EF, pred_EF)
ppc_bars_grouped(obs_EF, pred_EF, group = cut(log(patient_data$lab_NT_proBNP[!data_badhealth$EF_missing]), 3))
```


